Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-28 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update draft according to lisa/moritz morning revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2902 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 89f3a2f64a6d5bb6acb9b16a7ac4d28fbd7dde2b

Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first

@@

\cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. 

@@

All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
are built on additive regression models.
Generally these provide a powerful toolbox for modeling particularly smooth effects.
However, in the case of modeling circular data 
additive effects conditional on linear covariates are not intuitive
due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation

@@

@@

Moreover, abrupt changes 
might not be detected properly owing to the enforced additive structure and selected model 
specification such as the form of linear combinations.

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests 
do not require a pre-specified link function or any other model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while 
forests allow for modeling smooth changes as well. Thus, a high level of 
flexibility is provided while a full distributional model is specified in each terminal 
node offering a wide range of inference. Moreover, as covariates are employed for splitting only and not
considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.

>>>>

Mises distribution which is also known as the ``circular normal distribution''.
%This is also the basis
%which early regression models for circular data are built on as they regress
%the parameter(s) of the von Mises distribution on covariates. 
One of the first

@@

\cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a 
restricted interval of length $2\pi$. 
%%%
%%%In order to restrict the circular response to an interval of length $2\,pi\
%%%\cite{Johnson+Wehrly:1978} refined this idea by linintroducing link functions transforming
%%%the linear predictor to the restricted interval of length $2\pi$. 
%\cite{Johnson+Wehrly:1978} took this idea one step further by
%plugging in a link function such that the values of the linear predictor are
%transformed to a restricted interval of length $2\pi$. 

@@

However, apart from potential difficulties in the optimization procedure of circular GLMs, 
additive effects and their interpretation are not intuitive for circular responses.
Due to the highly non-linear link function, the representation of smooth transitions on the unit circle
is not straightforward; additionally, the same rotation

@@

%%%All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
%%%are built on additive regression models.
%%%Generally these provide a powerful toolbox for modeling particularly smooth effects.
%%%however, in the case of modeling circular data 
%%%additive effects conditional on linear covariates are not intuitive
%%%due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
%%%can be obtained in either positive or negative direction on the circle.

@@

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests 
do not require a pre-specified link function or any other model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while 
forests allow for modeling smooth changes as well. Thus, a high level of 
flexibility is provided while a full distributional model is specified in each terminal 
node offering a wide range of inference. Moreover, as covariates are employed for splitting only and not
considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.
Moreover, abrupt changes 
might not be detected properly owing to the enforced additive structure and selected model 
specification such as the form of linear combinations.
